---
title: "Analysis of genes with bootstrap and permutation (RD)"
author: "Yaromir Kobikov <kobikov.yaromir@gmail.com>"
date: "Date: `r format(Sys.time(), '%d/%m/%Y')`"
output:
 word_document: default
 html_document:
  toc: true
  toc_float:
   collapsed: false
editor_options:
 chunk_output_type: inline
---

ToDo: Codes to accompany paper ...put BioRxiv reference for the paper... 

```{r setup, include=FALSE}
source("GenesAnalysis.R")
```
# 
<!-- # https://thomasleeper.com/Rcourse/Tutorials/permutationtests.html -->

# Data load

ToDo: briefly explain what is in each of the three files
Hypoth [SO, UBI];
Spinal cord [SO, UBI, Control]

```{r data_load, include=FALSE}
main_dir <- "." # path to the base folder, containing data and results
path_rds <- "rds" # folder to store the results
DESIGN <- "RD" # only reverse design data will be used

dir.create(file.path(main_dir, path_rds), showWarnings = FALSE)

data_Hypoth <-
  file.path(main_dir,"data/Hypoth_SO_UBI.xlsx") %>% 
  read_excel(.name_repair = "universal", na = c("", "NA", "NULL")) %>% 
  mutate(Segment = "Hyp",
         Trt = recode_factor(Operation.type, SO = 'Sh', cSMC = "UBI"),
         Module = recode_factor(Meas..Side, L = 'lm', R = 'rm')) %>% 
  select(-Meas..Side, -Operation.type) %>% 
  relocate(Rat.ID, Segment, Trt, Module) %>% 
  pivot_wider(names_from = c("Module", "Segment"),
              values_from = !c("Rat.ID", "Module", "Segment", "Trt")) %>% 
  mutate(Design = factor("RD")) #####

data_SpinalC <-
  file.path(main_dir,"data/SpinalC_SO_UBI_Ctrl_RD_DD.xlsx") %>%
  read_excel(sheet = "Sheet2", .name_repair = "universal", na = c("", "NA", "NULL")) %>% 
  # filter(Design == "RD") %>%  #####
  mutate(Segment = "SpC",
         Trt = recode_factor(Operation.type, SO = 'Sh', cSMC = "UBI", SMC = "UBI"),
         Module = recode_factor(Meas..Side, L = 'lm', R = 'rm'),
         Design = factor(Design, c("RD", "DD")) #####
         )

## ToDo: if we print something, there should be at least a brief explanation. For example:
## Sanity check: If the file is correct, this should give
##   L    R <NA> 
##  86   40   20 
## How come measurement side is NA? what do we do with such data?
table(data_SpinalC$Operation.side, useNA = "ifany")

data_SpinalC <- 
  data_SpinalC %>% 
  select(-Meas..Side, -Operation.type, -Operation.side) %>% 
  relocate(Rat.ID, Segment, Trt, Module) %>% 
  pivot_wider(names_from = c("Module", "Segment"),
              values_from = !c("Rat.ID", "Module", "Segment", "Trt", "Design")) #####

data_groups_all <-
  file.path(main_dir,"data/groups.xlsx") %>%
  read_excel(sheet = "all_groups", .name_repair = "universal", na = c("", "NA", "NULL"))
```

Join measurements for Hypothalamus and Spinal Cord

```{r join_data_and_filtering, include=FALSE}
data_S.H <- 
  full_join(data_Hypoth, data_SpinalC, by=c("Rat.ID", "Trt", "Design")) %>%
  relocate(Rat.ID, Design) %>% 
  filter(Design == DESIGN) %>% # RD or DD design (2 reports)
  mutate(Trt = factor(Trt)) %>% 
  select(-Design)
table(data_S.H$Rat.ID, data_S.H$Trt, useNA = "ifany")
```

ToDo: A short explanation, for example:

To generate null distribution we permute data 
 - across rat IDs, which will also permute across Treatment
 - across Module (measurements on the left vs. right side of the segment) within each individual rat
 - across Segment (Hypothalamus vs. Spinal Cord) within each Module

```{r permutation}
permutation <- function(df) {
  
  d.t <- df %>%
    mutate(Trt = sample(Trt)) %>% # permute rat ids
    pivot_longer(cols = !c("Rat.ID", "Trt"),
                 names_sep = "_",
                 names_to = c("Gene", "Module", "Segment"),
                 values_to = "value") %>%
    group_by(Rat.ID, Gene, Segment) %>%
    mutate(Module = sample(Module)) %>% # permute left vs. right measurement sides
    ungroup %>%
    pivot_wider(names_from = Module,
                values_from = value) %>%
    group_by(Rat.ID, Gene) %>%
    mutate(Segment = sample(Segment)) %>% # permute Hypothalamus vs. Spinal Cord
    ungroup %>%
    pivot_longer(cols = c("lm", "rm"),
                 names_to = "Module",
                 values_to = "value") %>%
    pivot_wider(names_from = c("Gene", "Module", "Segment"),
                values_from = !c("Rat.ID", "Gene", "Module", "Segment", "Trt"))
  
  return(d.t[ , order(names(d.t))])
}
```

ToDo: brief description of the function, for example:

A function for boot() to compute correlations and test statistics for each group.
The first call is special, it should use observed data instead of permuted data.
Setting Verbose=TRUE will generate plots.

```{r correlation_core_for_bootstrapping}
corByGroup <- function(df, indices, dg, verbose=FALSE) {
  if(!all(indices == 1:nrow(df))) { df <- permutation(df) }

  df <- df %>%
    pivot_longer(cols = !c("Rat.ID", "Trt"),
                 names_sep = "_",
                 names_to = c("Gene", "Module", "Segment"),
                 values_to = "value")

  # for each gene, compute if it belongs to L or R network
  d.net <- df %>%
    group_by(Segment) %>%
    pivot_wider(names_from = Module, values_from = value) %>%
    # drop_na() %>%
    group_by(Gene, .add=TRUE) %>%
    summarise(Net = ifelse(mean(log2(lm/rm), na.rm=TRUE) > 0, "L", "R") , .groups = "drop_last")

  df <- left_join(df, d.net, by=c("Segment", "Gene"))

  d.t1_cor <- df %>%
    pivot_wider(names_from = c("Segment", "Trt", "Module", "Net", "Gene"),
                values_from = value) %>%
    select(-Rat.ID) %>%
    as.matrix %>%
    cor(method = "spearman", use = "pairwise.complete.obs")

  diag(d.t1_cor) <- NA

  res <- rep(NA_real_, 6*nrow(dg))

  for (i in 1:nrow(dg)) {

    ind1.1 <- grepl(pattern = dg[i, ]$CorrMatr1.rows, x = rownames(d.t1_cor))
    ind1.2 <- grepl(pattern = dg[i, ]$CorrMatr1.columns, x = colnames(d.t1_cor))

    ind2.1 <- grepl(pattern = dg[i, ]$CorrMatr2.rows, x = rownames(d.t1_cor))
    ind2.2 <- grepl(pattern = dg[i, ]$CorrMatr2.columns, x = colnames(d.t1_cor))

    d1 <- d.t1_cor[ind1.1, ind1.2]
    d2 <- d.t1_cor[ind2.1, ind2.2]
    
    ## network correlation strength = magnitude of correlations, averaged across all parwise correlations
    ## For symmetric correlation matrix each correlation is used two times, but it does not affect the average.
    res[i*6-5] <- mean(abs(d1), na.rm = TRUE)
    res[i*6-4] <- mean(abs(d2), na.rm = TRUE)
    res[i*6-3] <- res[i*6-5] - res[i*6-4]
    
    # res[i*6-2] <- mean(d1 > 0, na.rm = TRUE)
    # res[i*6-1] <- mean(d2 > 0, na.rm = TRUE)
    
    ## proportion of positive correlations in the network
    pos <- mean(d1 > 0, na.rm = TRUE)
    neg <- mean(d1 < 0, na.rm = TRUE)
    res[i*6-2] <- pos / (pos + neg)
    
    pos <- mean(d2 > 0, na.rm = TRUE)
    neg <- mean(d2 < 0, na.rm = TRUE)
    res[i*6-1] <- pos / (pos + neg)

    res[i*6] <- res[i*6-2] - res[i*6-1]
  }
  return(res)
}
```

```{r check, include=FALSE}
# Sanity check: columns of joint data are sorted
colnames(data_S.H[ , order(names(data_S.H))]) == colnames(permutation(data_S.H[ , order(names(data_S.H))]))
# Sanity check: results for the first test
corByGroup(df = data_S.H, indices = 1:nrow(data_S.H), dg = data_groups_all %>% slice(1:1))
```

# Bootstrap with permutation

```{r bootstrap}
#set.seed(1244070722)
## to avoid question "Why do you choose this random seed?", I usually used the date
set.seed(20220710)

data_S.H <- data_S.H[ , order(names(data_S.H))]
data_group <- data_groups_all

R = 1e6 # 1e6 permutations took about 18 hours using 10 CPUs

rds_file <- file.path(main_dir, path_rds, paste0(
  "all_groups_boot_", formatter(R), "_", my.cores, "_cores_", DESIGN, ".rds"
))

parallel <- ifelse(Sys.info()["sysname"] == "Linux", "multicore", "snow")
if(!file.exists(rds_file)) {
  start_time <- Sys.time()
  res_boot <- boot(data = data_S.H,
                 R = R,
                 statistic = corByGroup,
                 dg = data_group,
                 parallel = parallel,
                 ncpus = my.cores)
  end_time <- Sys.time()
  print(end_time - start_time)
  
  saveRDS(object = res_boot, file = rds_file)
} else {
  res_boot <- readRDS(rds_file)
}
```

```{r test_by_index}
index <- 8

print(plot(res_boot, index=index))
print(res_boot[["t0"]][index])
print(boot.ci(res_boot, index=index, type = c("perc"), conf = 0.95))

print(boot.pval(res_boot, type="perc", pval_precision = 1/R, index=index, theta_null = res_boot[["t0"]][index]))

# a <- boot.ci(res_boot, index=3, type = c("norm", "perc"), conf = 0.95)
```

```{r}
sessionInfo()
```
