---
title: "Analysis of genes with bootstrap and permutation (RD) Net [Sh, UBI] median"
author: "Yaromir Kobikov <kobikov.yaromir@gmail.com>"
date: "Date: `r format(Sys.time(), '%d/%m/%Y')`"
output:
 word_document: default
 html_document:
  toc: true
  toc_float:
   collapsed: false
editor_options:
 chunk_output_type: inline
---

ToDo: Codes to accompany paper ...put BioRxiv reference for the paper... 

```{r setup, include=FALSE}
source("GenesAnalysis.R")
```
# 
<!-- # https://thomasleeper.com/Rcourse/Tutorials/permutationtests.html -->

# Data load

ToDo: briefly explain what is in each of the three files
Hypothalamus [SO, UBI];
Spinal cord [SO, UBI, Control]

```{r data_load, include=FALSE}
main_dir <- "." # path to the base folder, containing data and results
path_rds <- "rds" # folder to store the results
DESIGN <- "RD" # only reverse design data will be used

dir.create(file.path(main_dir, path_rds), showWarnings = FALSE)

data_Hypoth <-
  file.path(main_dir,"data/Hypoth_SO_UBI.xlsx") %>% 
  read_excel(.name_repair = "universal", na = c("", "NA", "NULL")) %>% 
  mutate(Segment = "Hyp",
         Trt = recode_factor(Operation.type, SO = 'Sh', cSMC = "UBI"),
         Module = recode_factor(Meas..Side, L = 'lm', R = 'rm')) %>% 
  select(-Meas..Side, -Operation.type) %>% 
  relocate(Rat.ID, Segment, Trt, Module) %>% 
  pivot_wider(names_from = c("Module", "Segment"),
              values_from = !c("Rat.ID", "Module", "Segment", "Trt")) %>% 
  mutate(Design = factor("RD")) #####

data_SpinalC <-
  file.path(main_dir,"data/SpinalC_SO_UBI_Ctrl_RD_DD.xlsx") %>%
  read_excel(sheet = "Sheet2", .name_repair = "universal", na = c("", "NA", "NULL")) %>% 
  # filter(Design == "RD") %>%  #####
  mutate(Segment = "SpC",
         Trt = recode_factor(Operation.type, SO = 'Sh', cSMC = "UBI", SMC = "UBI"),
         Module = recode_factor(Meas..Side, L = 'lm', R = 'rm'),
         Design = factor(Design, c("RD", "DD")) #####
         )

## ToDo: if we print something, there should be at least a brief explanation. For example:
## Sanity check: If the file is correct, this should give
##   L    R <NA> 
##  86   40   20 
## How come measurement side is NA? what do we do with such data?
table(data_SpinalC$Operation.side, useNA = "ifany")

data_SpinalC <- 
  data_SpinalC %>% 
  select(-Meas..Side, -Operation.type, -Operation.side) %>% 
  relocate(Rat.ID, Segment, Trt, Module) %>% 
  pivot_wider(names_from = c("Module", "Segment"),
              values_from = !c("Rat.ID", "Module", "Segment", "Trt", "Design")) #####

data_groups_all <-
  file.path(main_dir,"data/groups.xlsx") %>%
  read_excel(sheet = "all_groups", .name_repair = "universal", na = c("", "NA", "NULL"))
```

Join measurements for Hypothalamus and Spinal Cord

```{r join_data_and_filtering, include=FALSE}
data_S.H <- 
  full_join(data_Hypoth, data_SpinalC, by=c("Rat.ID", "Trt", "Design")) %>%
  relocate(Rat.ID, Design) %>% 
  filter(Design == DESIGN) %>% # RD or DD design (2 reports)
  mutate(Trt = factor(Trt)) %>% 
  select(-Design)
table(data_S.H$Rat.ID, data_S.H$Trt, useNA = "ifany")
```

ToDo: A short explanation, for example:

To generate null distribution we permute data 
 - across rat IDs, which will also permute across Treatment
 - across Module (measurements on the left vs. right side of the segment) within each individual rat
 - across Segment (Hypothalamus vs. Spinal Cord) within each Module

```{r permutation}
permutation <- function(df) {
  
  d.t <- df %>%
    mutate(Trt = sample(Trt)) %>% # permute rat ids
    pivot_longer(cols = !c("Rat.ID", "Trt"),
                 names_sep = "_",
                 names_to = c("Gene", "Module", "Segment"),
                 values_to = "value") %>%
    group_by(Rat.ID, Gene, Segment) %>%
    mutate(Module = sample(Module)) %>% # permute left vs. right measurement sides
    ungroup %>%
    pivot_wider(names_from = Module,
                values_from = value) %>%
    group_by(Rat.ID, Gene) %>%
    mutate(Segment = sample(Segment)) %>% # permute Hypothalamus vs. Spinal Cord
    ungroup %>%
    pivot_longer(cols = c("lm", "rm"),
                 names_to = "Module",
                 values_to = "value") %>%
    pivot_wider(names_from = c("Gene", "Module", "Segment"),
                values_from = !c("Rat.ID", "Gene", "Module", "Segment", "Trt"))
  
  return(d.t[ , order(names(d.t))])
}
```

ToDo: brief description of the function, for example:

A function for boot() to compute correlations and test statistics for each group.
The first call is special, it should use observed data instead of permuted data.
Setting Verbose=TRUE will generate plots.

```{r correlation_core_for_bootstrapping}
corByGroup <- function(df, indices, dg, verbose=FALSE) {
  if(!all(indices == 1:nrow(df))) { df <- permutation(df) }
  
  if (verbose) { buffer <- list() }

  df <- df %>%
    pivot_longer(cols = !c("Rat.ID", "Trt"),
                 names_sep = "_",
                 names_to = c("Gene", "Module", "Segment"),
                 values_to = "value")

  # for each gene, compute if it belongs to L or R network
  d.net <- df %>%
    group_by(Segment) %>%
    pivot_wider(names_from = Module, values_from = value) %>%
    # drop_na() %>%
    group_by(Gene, .add=TRUE) %>%
    summarise(Net_value = median(log2(lm/rm), na.rm=TRUE), # median or mean 
              Net = ifelse(Net_value > 0, "L", "R") , .groups = "drop_last")
  
  # collect data for report
  if (verbose) { buffer[['Networks']] <- d.net }

  df <- left_join(df, d.net, by=c("Segment", "Gene")) %>% select(-Net_value)

  d.t1_cor <- df %>%
    pivot_wider(names_from = c("Segment", "Trt", "Module", "Net", "Gene"),
                values_from = value) %>%
    select(-Rat.ID) %>%
    as.matrix %>%
    cor(method = "spearman", use = "pairwise.complete.obs")

  diag(d.t1_cor) <- NA

  res <- rep(NA_real_, 6*nrow(dg))

  for (i in 1:nrow(dg)) {

    ind1.1 <- grepl(pattern = dg[i, ]$CorrMatr1.rows, x = rownames(d.t1_cor))
    ind1.2 <- grepl(pattern = dg[i, ]$CorrMatr1.columns, x = colnames(d.t1_cor))

    ind2.1 <- grepl(pattern = dg[i, ]$CorrMatr2.rows, x = rownames(d.t1_cor))
    ind2.2 <- grepl(pattern = dg[i, ]$CorrMatr2.columns, x = colnames(d.t1_cor))

    d1 <- d.t1_cor[ind1.1, ind1.2]
    d2 <- d.t1_cor[ind2.1, ind2.2]
    
    # collect data for report
    if (verbose) {
      buffer[[as.character(i+1)]] <- list(
        description = dg[i, ]$Description,
        matr1 = d1,
        matr2 = d2,
        matr1.name.rows = dg[i, ]$CorrMatr1.rows,
        matr1.name.cols = dg[i, ]$CorrMatr1.columns,
        matr2.name.rows = dg[i, ]$CorrMatr2.rows,
        matr2.name.cols = dg[i, ]$CorrMatr2.columns
      )
    }
    
    ## network correlation strength = magnitude of correlations, averaged across all pairwise correlations
    ## For symmetric correlation matrix each correlation is used two times, but it does not affect the average.
    res[i*6-5] <- mean(abs(d1), na.rm = TRUE)
    res[i*6-4] <- mean(abs(d2), na.rm = TRUE)
    res[i*6-3] <- res[i*6-5] - res[i*6-4]
    
    # res[i*6-2] <- mean(d1 > 0, na.rm = TRUE)
    # res[i*6-1] <- mean(d2 > 0, na.rm = TRUE)
    
    ## proportion of positive correlations in the network
    pos <- mean(d1 > 0, na.rm = TRUE)
    neg <- mean(d1 < 0, na.rm = TRUE)
    res[i*6-2] <- pos / (pos + neg)
    
    pos <- mean(d2 > 0, na.rm = TRUE)
    neg <- mean(d2 < 0, na.rm = TRUE)
    res[i*6-1] <- pos / (pos + neg)

    res[i*6] <- res[i*6-2] - res[i*6-1]
  }
  if (!verbose) return(res)
  else return(list(data.viz=buffer, result=res))
}
```

```{r check, include=FALSE}
# Sanity check: columns of joint data are sorted
colnames(data_S.H[ , order(names(data_S.H))]) == colnames(permutation(data_S.H[ , order(names(data_S.H))]))
# Sanity check: results for the first test
corByGroup(df = data_S.H, indices = 1:nrow(data_S.H), dg = data_groups_all %>% slice(1:1))
```

# Bootstrap with permutation

```{r bootstrap}
# To reproduce results during parallel operation. 
# Only for parallel = "multicore" (not available on Windows).
RNGkind("L'Ecuyer-CMRG") # from https://cran.r-project.org/web/packages/boot/boot.pdf (14 page)
set.seed(20220710)

data_S.H <- data_S.H[ , order(names(data_S.H))]
data_group <- data_groups_all

R = 1e3 # 1e6 permutations took about 18 hours using 10 CPUs

rds_file <- file.path(main_dir, path_rds, paste0(
  "all_groups_boot_", formatter(R), "_", my.cores, "_cores_", DESIGN, "_Net_Sh_UBI_median.rds"
))

parallel <- ifelse(Sys.info()["sysname"] == "Linux", "multicore", "snow")
if(!file.exists(rds_file)) {
  start_time <- Sys.time()
  res_boot <- boot(data = data_S.H,
                 R = R,
                 statistic = corByGroup,
                 dg = data_group,
                 parallel = parallel,
                 ncpus = my.cores)
  end_time <- Sys.time()
  print(end_time - start_time)
  
  saveRDS(object = res_boot, file = rds_file)
} else {
  res_boot <- readRDS(rds_file)
}
```

```{r test_by_index}
index <- 8

# print(plot(res_boot, index=index)) # At 1e6 it takes more than 20 GB of RAM, be careful.
print(res_boot[["t0"]][index])
print(boot.ci(res_boot, index=index, type = c("perc"), conf = 0.95))

print(boot.pval(res_boot, type="perc", pval_precision = 1/R, index=index, theta_null = res_boot[["t0"]][index]))

# a <- boot.ci(res_boot, index=3, type = c("norm", "perc"), conf = 0.95)
```

```{r CI_by_groups, include=FALSE}

groups_results <- paste0("./data/Net_Sh_UBI_median_groups_result_", format(Sys.time(), '%d_%m_%Y'), ".xlsx")

if(!file.exists(groups_results)) {
  tmp <- rep(NA_real_, 6*nrow(data_groups_all))

  conf <- 0.95
  
  for (i in 1:nrow(data_groups_all)) {
    index <- i*6
    
    ci <- boot.ci(res_boot, index=index - 3, type = c("perc"), conf = conf)
    tmp[i*6-5] <- ci$percent[4]
    tmp[i*6-4] <- ci$percent[5]
    tmp[i*6-3] <- boot.pval(res_boot, type="perc", pval_precision = 1/R, index=index - 3, theta_null = res_boot[["t0"]][index - 3])

    ci <- boot.ci(res_boot, index=index, type = c("perc"), conf = conf)
    tmp[i*6-2] <- ci$percent[4]
    tmp[i*6-1] <- ci$percent[5]
    tmp[i*6] <- boot.pval(res_boot, type="perc", pval_precision = 1/R, index=index, theta_null = res_boot[["t0"]][index])
  }
  
  for (index in 1:6) {
    data_groups_all[index + 7] <- res_boot[["t0"]][seq(index, length(res_boot[["t0"]]), 6)]
    data_groups_all[index + 13] <- tmp[seq(index, length(res_boot[["t0"]]), 6)]
  }


  data_groups_all_sorted <-
    data_groups_all %>%
    filter(analyze == "yes") %>%
    select(-analyze) %>%
    group_by(Segment) %>%
    mutate(BH.pStr = p.adjust(pval.CorStr, method = "BH"),
           BH.pProp = p.adjust(pval.CorProp, method = "BH"))
  
  write.xlsx(data_groups_all_sorted, groups_results)
} else {
  data_groups_all_sorted <-
    groups_results %>%
    read_excel(.name_repair = "universal", na = c("", "NA", "NULL"))
}
  
```

# Visualizing data by groups

```{r list_data_by_groups, include=FALSE}
groups <- data_groups_all %>% filter(analyze == 'yes')
data.groups.list <- corByGroup(df = data_S.H[ , order(names(data_S.H))], indices = 1:nrow(data_S.H[ , order(names(data_S.H))]), dg = groups, verbose = TRUE)
```

```{r custome_theme, include=FALSE}
my.theme <- function(gg, 
                          title="", 
                          labs_x="",
                          labs_y="",
                          margin_list=c(0,0,0,0),
                          title.x.hjust=1,
                          title.y.hjust=0.5,
                          text.x.angle=0) {
  return(
    gg +
      labs(fill = "", title = title, x=labs_x, y=labs_y) +
      theme(
        legend.position = "none",
        
        plot.title = element_text(size=14),
        plot.margin = margin(margin_list[1], margin_list[2], margin_list[3], margin_list[4], "cm"),
        
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.55),
        panel.background = element_blank(),
        
        axis.ticks = element_line(size = 0.55, colour="black"),
        axis.ticks.length.x = unit(.13, "inches"),
        axis.ticks.length.y = unit(.13, "inches"),
        
        axis.title.x = element_text(size = 12, family = "Arial", hjust = title.x.hjust),
        axis.title.y = element_text(size = 12, family = "Arial", hjust = title.y.hjust),
        
        axis.text = element_text(size=12, colour="black"),
        axis.text.x = element_text(angle=text.x.angle, size=12, family = "Arial", hjust=0.5, margin = margin(t = 0.35, unit = "cm")),
        axis.text.y = element_text(angle=0, size=12, family = "Arial", margin = margin(r = 0.2, l=0.2, unit = "cm"))
      ) +
      scale_fill_manual(values = my.colors)
  )
}
```

## (Bar) Plots of Networks by Hypothalamus and Spinal Cord

```{r networks_bar_plots_hyp_spc, fig.width=6, fig.height=6}
data <- data.groups.list$data.viz$Networks

for (item in unique(data$Segment)) {
  d.net_ <- data %>%
    filter(Segment == item) %>% 
    mutate(Gene = fct_reorder(Gene, desc(Net_value)))
  name <- paste0(item, " Networks Sh UBI median")
  plt[[name]] <- (ggplot(d.net_, aes(x=fct_rev(Gene), y=Net_value, fill=Net)) +
                    geom_bar(stat='identity', width=.5) +
                    coord_flip()) %>% my.theme
  print(plt[[name]])
}
```

## Plots for each groups
Heatmap, Violin, Proportions

```{r custome_theme_heatmap, include=FALSE}
my.theme.heatmap <- function(gg) {
  return(gg +
           scale_fill_gradientn(colours = heatmap.colors, limits = c(-1, 1), breaks = c(-1, 0, 1), expand = c(0, 0)) +
           theme(legend.position = "top", 
                 legend.box = "vertical",
                 
                 plot.title = element_text(size=14),
                 plot.margin = margin(0, 0, 0, 0, "cm"),
                 
                 panel.grid.major = element_blank(), 
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank(),
                 
                 axis.ticks = element_blank(),
                 
                 axis.text = element_text(size=12, colour="black"),
                 axis.text.x = element_text(angle=90, size=12),
                 axis.text.y = element_text(angle=0, size=12, family = "", margin = margin(r=0.2, l=0.2, unit = "cm"))
           )
  )
}
```

```{r heatmap_fun, include=FALSE}
get_heatmap <- function(matrix, x.name, y.name) {
  d <- matrix %>% as.data.frame() %>% rownames_to_column("rows")
  
  if (x.name == y.name) {
    tmp_names <- d$rows
    d <- subset(d, select = -rows)
    d[lower.tri(d)] <- NA
    d$rows <- tmp_names
  }
  
  d <- d %>%
    pivot_longer(-c(rows), names_to = "cols", values_to = "Correlation coefficient") %>% 
    drop_na(.)
  
  return ((d %>% 
    ggplot(aes(fct_rev(cols), rows)) + 
    geom_tile(aes(fill = `Correlation coefficient`), width=1)) %>% 
    my.theme.heatmap)
}
```

```{r violin_proportions_plots, include=FALSE}
get_violin_prop_plots <- function(matrx.1, matrx.2, x.name.1, y.name.1, x.name.2, y.name.2) {
  d1 <- matrx.1 %>% as.data.frame() %>% rownames_to_column('rows')
  d1 <- d1 %>%
    pivot_longer(-c(rows), names_to = "cols", values_to = "Rho") %>% 
    drop_na(.) %>%
    mutate(groups = paste0(x.name.1, ', ', y.name.1), 
           sign = ifelse(Rho > 0, 'pos', 'neg'))
  
  d2 <- matrx.2 %>% as.data.frame() %>% rownames_to_column("rows")
  d2 <- d2 %>%
    pivot_longer(-c(rows), names_to = "cols", values_to = "Rho") %>%
    drop_na(.) %>%
    mutate(groups = paste0(x.name.2, ', ', y.name.2), 
           sign = ifelse(Rho > 0, 'pos', 'neg'))
  
  dd <- rbind(d1, d2)
  
  gg1 <- (dd %>% 
    mutate(Rho = abs(Rho)) %>% 
    ggplot(aes(x=groups, y=Rho, fill=groups)) +
    geom_violin() +
    stat_summary(fun="median", geom = "crossbar", width=0.5) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = .25))) %>% 
    my.theme(labs_y = '|Rho|')
  
  gg2 <- (dd %>%
            group_by(groups, sign) %>%
            summarise(n=n(), .groups="drop") %>%
            ggplot(aes(fill=sign, y=n, x=groups)) + 
            geom_bar(position="fill", stat="identity", width=.5) +
            geom_hline(yintercept = 0.5, linetype="dashed")) %>% 
    my.theme
  
  return(list(gg1, gg2))
}
```


```{r plots_for_each_groups, fig.width=6, fig.height=6}
data <- data.groups.list$data.viz
ds <- data[2:length(data)]

print.it <- FALSE

help.print.fun <- function(name, plt) {
  print(name)
  print(plt)
}

for (i in names(ds)) {
  
  name <- paste0('Heatmap, ', i, ', ', ds[[i]]$description, ' Gr1, rows: ', ds[[i]]$matr1.name.rows, ', cols: ', ds[[i]]$matr1.name.cols)
  plt[[name]] <- get_heatmap(ds[[i]]$matr1, ds[[i]]$matr1.name.rows, ds[[i]]$matr1.name.cols)
  if (print.it) help.print.fun(name, plt[[name]])

  name <- paste0('Heatmap, ', i, ', ', ds[[i]]$description, ' Gr2, rows: ', ds[[i]]$matr2.name.rows, ', cols: ', ds[[i]]$matr2.name.cols)
  plt[[name]] <- get_heatmap(ds[[i]]$matr2, ds[[i]]$matr2.name.rows, ds[[i]]$matr2.name.cols)
  if (print.it) help.print.fun(name, plt[[name]])

  tmp <- get_violin_prop_plots(ds[[i]]$matr1, ds[[i]]$matr2,
                               ds[[i]]$matr1.name.rows, ds[[i]]$matr1.name.cols,
                               ds[[i]]$matr2.name.rows, ds[[i]]$matr2.name.cols)

  name <- paste0(i, ', ', ds[[i]]$description,
           ', G1 rows: ', ds[[i]]$matr1.name.rows, ', cols: ', ds[[i]]$matr1.name.cols,
           ', G2 rows: ', ds[[i]]$matr2.name.rows, ', cols: ', ds[[i]]$matr2.name.cols)
  
  name.violin <- paste0('Violin, ', name)
  plt[[name.violin]] <- tmp[[1]]
  if (print.it) help.print.fun(name.violin, plt[[name.violin]])

  name.proportion <- paste0('Proportions, ', name)
  plt[[name.proportion]] <- tmp[[2]]
  if (print.it) help.print.fun(name.proportion, plt[[name.proportion]])
}
```

# Editable PowerPoint plots generated...

```{r pptx, message=TRUE, warning=TRUE, include=FALSE, results='hide'}

myname <- "corr_groups_Net_Sh_UBI_median_"

doc <- read_pptx()

for(nm in names(plt)) {
  doc <- doc %>%
    add_slide(layout = "Title and Content", master = "Office Theme") %>%
    ph_with(value = rvg::dml(ggobj = plt[[nm]]),
          location = ph_location(width=6, height=6),
          bg = "transparent" ) %>%
    ph_with(value = nm,
          location = ph_location_type(type = "title") )
}

doc %>% print(target = paste0("FIGs_", myname, format(Sys.time(), '%d_%m_%Y'), ".pptx"))

```

```{r session_info}
sessionInfo()
```
